---
title: "Forecasting Traffic at Sensor 198"
author: "Ana Blanco Lara, Imen Rabouhi Announ, Xavier Dubrul & Camille Toan"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tsibble)
library(lubridate)
library(fable)
library(feasts)
library(readr)
library(zoo)
library(tidyr)
```

# Step 1: Exploration and Visualization

```{r}
# Load and preprocess the traffic data
df_traffic <- read_csv("traffic_data.csv")[-1, ]  # Remove header row
df_traffic <- df_traffic %>%
  mutate(
    datetime = dmy_hm(Time),
    traffic = as.numeric(Sensor198)
  ) %>%
  select(datetime, traffic) %>%
  as_tsibble(index = datetime)
```

## Time Series Overview

```{r}
df_traffic %>% autoplot(traffic) +
  labs(title = "Traffic at Sensor 198", x = "Time", y = "Traffic")
```

## Traffic Distribution

```{r}
ggplot(df_traffic, aes(x = traffic)) +
  geom_histogram(binwidth = 0.05, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Traffic Values", x = "Traffic", y = "Count")
```

## Seasonality (Monthly, Weekly)

```{r}
gg_season(df_traffic, traffic, period = "month") +
  labs(title = "Monthly Seasonality", x = "Date", y = "Traffic")

gg_season(df_traffic, traffic, period = "week") +
  labs(title = "Weekly Seasonality", x = "Date", y = "Traffic")
```

## Dailly Pattern

```{r}
df_traffic %>%
  mutate(weekday = factor(weekdays(datetime), levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")),
         hour = hour(datetime)) %>%
  group_by(weekday, hour) %>%
  summarise(mean_traffic = mean(traffic, na.rm = TRUE)) %>%
  ggplot(aes(x = hour, y = mean_traffic)) +
  geom_line() +
  facet_wrap(~ weekday, ncol = 3) +
  labs(title = "Average Hourly Traffic by Day", x = "Hour", y = "Traffic")
```

## Autocorrelation

```{r}
df_traffic %>% ACF(traffic) %>% autoplot() +
  labs(title = "Autocorrelation Function")
```

## Holiday Impact

```{r}
# Add holiday flag
df_traffic <- df_traffic %>%
  mutate(date = as_date(datetime))

holidays <- as_date(c(
  "2015-01-01", "2015-01-19", "2015-02-16", "2015-03-31", "2015-05-25",
  "2015-07-04", "2015-09-07", "2015-10-12", "2015-11-11", "2015-11-26",
  "2015-11-27", "2015-12-25", "2016-01-01", "2016-01-18", "2016-02-15",
  "2016-03-31", "2016-05-30", "2016-07-04", "2016-09-05", "2016-10-10",
  "2016-11-11", "2016-11-24", "2016-11-25", "2016-12-25"
))

df_traffic <- df_traffic %>% mutate(holiday = as.integer(date %in% holidays))

df_traffic %>%
  group_by(holiday) %>%
  summarise(mean_traffic = mean(traffic, na.rm = TRUE)) %>%
  mutate(holiday = factor(holiday, labels = c("Non-Holiday", "Holiday"))) %>%
  ggplot(aes(x = holiday, y = mean_traffic)) +
  geom_col(fill = "lightblue") +
  labs(title = "Mean Traffic by Holiday Status", x = "Day Type", y = "Mean Traffic")
```

## Weather Impact

```{r}
weather <- read_csv("~/Desktop/weather_data.csv") %>%  # Replace with your actual filename
  mutate(
    date = as_date(`DATE`),  # Replace `DATE` with actual column name
    precipitation = as.numeric(`PRCP`),  # Replace with actual precipitation column
    snow = as.numeric(`SNOW`)            # Replace with actual snow column
  ) %>%
  select(date, precipitation, snow) %>%
  drop_na()

df_traffic <- df_traffic %>%
  mutate(date = as_date(datetime))

df_traffic <- weather %>%
  left_join(df_traffic, by = "date")

df_traffic <- df_traffic %>%
  select(datetime, date, traffic, holiday, precipitation, snow)
```

```{r}
# Effect of precipitation on traffic
ggplot(df_traffic, aes(x = precipitation, y = traffic)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "loess", color = "blue") +
  labs(title = "Traffic vs Precipitation", x = "Precipitation (inches)", y = "Traffic")

# Effect of snow on traffic
ggplot(df_traffic, aes(x = snow, y = traffic)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "loess", color = "steelblue") +
  labs(title = "Traffic vs Snowfall", x = "Snowfall (inches)", y = "Traffic")
```

# Step 2: Cleaning and Wrangling

```{r}
# 1. Fix duplicate datetime entries (average values per hour if needed)
df_clean <- df_traffic %>%
  group_by(datetime) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE), .groups = "drop")

# 2. Convert to a tsibble and fill missing hourly timestamps
df_clean <- df_clean %>%
  as_tsibble(index = datetime) %>%
  fill_gaps()

# 3. Identify and replace full "zero-traffic" days with NA
zero_days <- df_clean %>%
  mutate(date = as_date(datetime)) %>%
  group_by(date) %>%
  summarise(daily_total = sum(traffic, na.rm = TRUE), .groups = "drop") %>%
  filter(daily_total == 0) %>%
  pull(date)

df_clean <- df_clean %>%
  mutate(date = as_date(datetime)) %>%
  mutate(traffic = if_else(date %in% zero_days, NA_real_, traffic)) %>%
  select(-date)

# 4. Interpolate missing values
df_clean <- df_clean %>%
  mutate(traffic = na.approx(traffic, na.rm = FALSE))

# 5. Detect outliers with STL decomposition
decomposition <- df_clean %>%
  model(STL(traffic ~ season(window = "periodic"), robust = TRUE)) %>%
  fabletools::components()

iqr <- IQR(decomposition$remainder, na.rm = TRUE)

outliers <- decomposition %>%
  filter(remainder > quantile(remainder, 0.75, na.rm = TRUE) + 3 * iqr |
         remainder < quantile(remainder, 0.25, na.rm = TRUE) - 3 * iqr)

decomposition %>% autoplot() +
  labs(title = "STL Decomposition of Cleaned Traffic Data") +
  theme_minimal()
```

# Step 3: Modeling

## 3.1 Train-Test Split

```{r}
# We'll use the last 744 hours of 2016 (i.e. January 1st 00:00 is excluded) as test set
train_data <- df_clean %>% filter(datetime < ymd_h("2016-12-01 00"))
test_data <- df_clean %>% filter(datetime >= ymd_h("2016-12-01 00"))
```

## 3.2 Model Comparison and Selection

```{r}
# Fit models on training data (for validation only)
fit_ets_train <- train_data %>% 
model(ETS(traffic ~ error("A") + trend("N") + season("N")))

fit_arima_train <- train_data %>% 
model(ARIMA(traffic))
```

```{r}
fit_arima_xreg_train <- train_data %>%
  model(ARIMA(traffic ~ holiday + precipitation + snow))

fit_arima_xreg_train_h <- train_data %>%
  model(ARIMA(traffic ~ holiday))

fit_arima_xreg_train_p <- train_data %>%
  model(ARIMA(traffic ~ precipitation))

fit_arima_xreg_train_s <- train_data %>%
  model(ARIMA(traffic ~ snow))

fit_arima_xreg_train_hp <- train_data %>%
  model(ARIMA(traffic ~ holiday + precipitation))

fit_arima_xreg_train_ps <- train_data %>%
  model(ARIMA(traffic ~ precipitation + snow))

fit_arima_xreg_train_hs <- train_data %>%
  model(ARIMA(traffic ~ holiday + snow))

fc_xreg <- list(
  full      = forecast(fit_arima_xreg_train, new_data = test_data),
  holiday   = forecast(fit_arima_xreg_train_h, new_data = test_data),
  precip    = forecast(fit_arima_xreg_train_p, new_data = test_data),
  snow      = forecast(fit_arima_xreg_train_s, new_data = test_data),
  hol_prec  = forecast(fit_arima_xreg_train_hp, new_data = test_data),
  prec_snow = forecast(fit_arima_xreg_train_ps, new_data = test_data),
  hol_snow  = forecast(fit_arima_xreg_train_hs, new_data = test_data)
)

acc_xreg <- bind_rows(
  accuracy(fc_xreg$full, test_data)      %>% mutate(model = "holiday + prec + snow"),
  accuracy(fc_xreg$holiday, test_data)   %>% mutate(model = "holiday"),
  accuracy(fc_xreg$precip, test_data)    %>% mutate(model = "precipitation"),
  accuracy(fc_xreg$snow, test_data)      %>% mutate(model = "snow"),
  accuracy(fc_xreg$hol_prec, test_data)  %>% mutate(model = "holiday + precip"),
  accuracy(fc_xreg$prec_snow, test_data) %>% mutate(model = "precip + snow"),
  accuracy(fc_xreg$hol_snow, test_data)  %>% mutate(model = "holiday + snow")
)

acc_xreg %>%
  select(model, RMSE, MAE, MASE) %>%
  arrange(RMSE)
```

```{r}
fit_tslm_train <- train_data %>%
  model(TSLM(traffic ~ trend() + season() + holiday + precipitation + snow))

fit_tslm_train_h <- train_data %>%
  model(TSLM(traffic ~ trend() + season() + holiday))

fit_tslm_train_p <- train_data %>%
  model(TSLM(traffic ~ trend() + season() + precipitation))

fit_tslm_train_s <- train_data %>%
  model(TSLM(traffic ~ trend() + season() + snow))

fit_tslm_train_hp <- train_data %>%
  model(TSLM(traffic ~ trend() + season() + holiday + precipitation))

fit_tslm_train_ps <- train_data %>%
  model(TSLM(traffic ~ trend() + season() + precipitation + snow))

fit_tslm_train_hs <- train_data %>%
  model(TSLM(traffic ~ trend() + season() + holiday  + snow))

fc_tslm <- list(
  full      = forecast(fit_tslm_train, new_data = test_data),
  holiday   = forecast(fit_tslm_train_h, new_data = test_data),
  precip    = forecast(fit_tslm_train_p, new_data = test_data),
  snow      = forecast(fit_tslm_train_s, new_data = test_data),
  hol_prec  = forecast(fit_tslm_train_hp, new_data = test_data),
  prec_snow = forecast(fit_tslm_train_ps, new_data = test_data),
  hol_snow  = forecast(fit_tslm_train_hs, new_data = test_data)
)

acc_tslm <- bind_rows(
  accuracy(fc_tslm$full, test_data)      %>% mutate(model = "holiday + prec + snow"),
  accuracy(fc_tslm$holiday, test_data)   %>% mutate(model = "holiday"),
  accuracy(fc_tslm$precip, test_data)    %>% mutate(model = "precipitation"),
  accuracy(fc_tslm$snow, test_data)      %>% mutate(model = "snow"),
  accuracy(fc_tslm$hol_prec, test_data)  %>% mutate(model = "holiday + precip"),
  accuracy(fc_tslm$prec_snow, test_data) %>% mutate(model = "precip + snow"),
  accuracy(fc_tslm$hol_snow, test_data)  %>% mutate(model = "holiday + snow")
)

acc_tslm %>%
  select(model, RMSE, MAE, MASE) %>%
  arrange(RMSE)
```

## 3.3 Model Accuracy

```{r}
# Forecast models on test data
fc_ets_val <- forecast(fit_ets_train, new_data = test_data)
fc_arima_val <- forecast(fit_arima_train, new_data = test_data)
fc_arima_xreg_val <- forecast(fit_arima_xreg_train_h, new_data = test_data)
fc_tslm_val <- forecast(fit_tslm_train_hp, new_data = test_data)

# Forecast accuracies summary
bind_rows(
  accuracy(fc_ets_val, test_data) %>% mutate(model = "ETS"),
  accuracy(fc_arima_val, test_data) %>% mutate(model = "ARIMA"),
  accuracy(fc_arima_xreg_val, test_data) %>% mutate(model = "ARIMA + XREG"),
  accuracy(fc_tslm_val, test_data) %>% mutate(model = "TSLM")
)
```

## 3.4 Model Fitting

```{r}
# ETS
fit_ets <- df_clean %>% model(ETS(traffic ~ error("A") + trend("N") + season("N")))

# ARIMA
fit_arima <- df_clean %>% model(ARIMA(traffic))

# ARIMA with weather + holiday
fit_arima_xreg <- df_clean %>%
  model(ARIMA(traffic ~ holiday))

# TSLM with weather + holiday
fit_tslm <- df_clean %>%
  model(TSLM(traffic ~ trend() + season() + holiday + precipitation))
```

## 3.5 Model Summaries

```{r}
report(fit_ets)
report(fit_arima)
report(fit_arima_xreg)
report(fit_tslm)
```

# Step 4: Forecasting and Validation

## 4.1 January 2017 Data Preparation

```{r}
# Generate January 2017 weather
jan_avg_weather <- weather %>%
  filter(month(date) == 1, year(date) %in% c(2015, 2016)) %>%
  summarise(
    precipitation = mean(precipitation, na.rm = TRUE),
    snow = mean(snow, na.rm = TRUE)
  )

# Generate January 2017 hourly timestamps
jan_2017 <- tibble(
  datetime = seq.POSIXt(
    from = as.POSIXct("2017-01-01 00:00:00", tz = "UTC"),
    to = as.POSIXct("2017-01-31 23:00:00", tz = "UTC"),
    by = "hour"
  )
) %>%
  mutate(date = as_date(datetime),
    holiday = as.integer(date %in% holidays),
    precipitation = jan_avg_weather$precipitation,
    snow = jan_avg_weather$snow)

# Generate January 2017 holiday
jan_holidays_2017 <- as_date(c("2017-01-02", "2017-01-16"))

jan_2017 <- jan_2017 %>%
  mutate(holiday = as.integer(date %in% jan_holidays_2017))

# Transform into tsibble
jan_2017 <- jan_2017 %>%
  as_tsibble(index = datetime)
```

## 4.2 Forecast - January 2017

```{r}
# Forecast horizon: 744 hours = entire January 2017
horizon <- "744 hours"

# ETS
fc_ets <- fit_ets %>% forecast(h = horizon)

# ARIMA
fc_arima <- fit_arima %>% forecast(h = horizon)

# ARIMA with external regressors
fc_arima_xreg_2017 <- fit_arima_xreg %>% forecast(new_data = jan_2017)

# TSLM with external regressors
fc_tslm_2017 <- fit_tslm %>% forecast(new_data = jan_2017)

# Visualize the forecasts
fc_ets %>%
  autoplot(df_clean) +
  labs(title = "ETS Forecast for January 2017",
       x = "Time", y = "Traffic")

fc_arima %>%
  autoplot(df_clean) +
  labs(title = "ARIMA Forecast for January 2017",
       x = "Time", y = "Traffic")

fc_arima_xreg_2017 %>%
  autoplot(df_clean) +
  labs(title = "ARIMA XREG Forecast for January 2017",
       x = "Time", y = "Traffic")

fc_tslm_2017 %>%
  autoplot(df_clean) +
  labs(title = "TSLM Forecast for January 2017",
       x = "Time", y = "Traffic")
```
